Changes in accessibility to foodbanks during COVID-19 and implications for the food security of vulnerable populations in Hamilton, Ontario

### 2021-04-20 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
First draft of methods section and minor edits in the introduction

It is known that the urban food environment, within which people make their daily food choices, is essential in influencing eating behaviours and health outcomes, based on factors such as food availability, ease of geographic accessibility and socio-demographic variations [@paez2010relative; @vanderlee2017food; @widener2018spatial]. However, while there is a wealth of literature that has examined the topic of geographic accessibility to healthy food through the "food desert" concept, there has been little research into accessibility to food banks. Although previous work has explored differences in _accessing_ food banks, such as how some households utilize food banks over short periods of time while others regularly utilize food banks as longer-term resource [e.g., @enns2020experiences], we are not aware of any research that has focused on the geographic component of accessibility.

@@

For this study, we aim to look at how the landscape of food banks and related services (e.g. low-cost or free meal service providers) available in Hamilton, Ontario, has changed before and during the pandemic. Have the number of open food bank services diminished? If so, what was the accessibility to foodbanks before and during the pandemic, from the perspective of low income households? And finally, who are most likely to have been impacted by changes to the accessibility landscape? This paper will first look at the distribution of food banks and related services before and during the pandemic. Then, we use the balanced floating catchment areas approach of Paez et al. [@paez2019demand] to investigate the accessibility situation. We use a fully disaggregated approach based on parcel-level data. Socio-economic and demographic data are drawn from the latest Census of Canada (2016), whereas travel information is from the most recent regional travel survey from 2016. This paper follows reproducible research recommendations [see @brunsdon2020opening], and the research was conducted using open source tools for transportation analysis [@lovelace2021open]. The code and data necessary to reproduce the analysis are available in a public repository^[add repository].

@@

Food insecurity is the inability to acquire and consume an adequate amount or good quality food, leading to inadequate nutrient intake (Kirkpatrick and Tarasuk 2008; Tarasuk and Vogt 2009). This nutrient deficiency has been causing major health concerns in Canadians, and particularly those who are at a socioeconomic disadvantage (Bazerghi, McKay, and Dunn 2016). Previous studies have aimed to look at the relationship between the built food environment and socio-demographic characteristics with qualitative and quantitative, or a mixed -method approach.

Quantitatively, official governmental surveys have been able to assist with data, such as the Household Food Security Survey Module (HFSSM), the Canadian Community Health Surveys (CCHS) , the Longitudinal and International Study of Adults (LISA), and official classifications determined by Health Canada in relation to  sociodemographic variables (El-hajj and Benhin 2021; Gundersen et al. 2018; Kirkpatrick and Tarasuk 2008; Tarasuk and Vogt 2009). Studies have also aimed to assess food availability of healthy foods (e.g., fruit and vegetables) at supermarkets in relation to sociodemographic characteristics and geographic accessibility (Latham and Moffat 2007). In terms of findings, studies have generated inconsistent relationships between their evaluated availability of food and sociodemographic characteristics that go in hand.

## Food Banks

The number of food banks has been increasing steadily in Canada (Wakefield et al. 2013). What was supposed to be a temporary solution to accommodate those in hunger due to job retrenchments and economic downfalls since the 1980s, has now developed into a community practice to secure emergency food supplies for those in need (Loopstra and Tarasuk 2012; Wakefield et al. 2013). Food banks were developed as a part of a community response to aide those with inadequate food by voluntarily giving meals and ingredients away (Loopstra and Tarasuk 2012; Riches 2002). The scope and objectives of food banks can vary by region and by country. These organizations can include not only prepared meals and aliments, but also shared spaces to connect in community gardens and community kitchens (Wakefield et al. 2013) or referred to as ‘food pantry’ and ‘food shelf’, where many frequent as their primary location to get food (Bazerghi et al. 2016). However, surveys revealed that only 20 to 30 percent of those experiencing food insecurity were found to frequent food banks in Canada (Tarasuk et al. 2014).

Previous studies also question if food banks are able to offer a rounded nutritious supply of food, and if food banks are a sustainable practice for those in need to continuously obtain their food from (Bazerghi et al. 2016; Riches 2002).

@@

Currently with the COVID-19 pandemic, disrupted economies, rising unemployment rates, and alarming poverty levels have disrupted the food environment by causing higher rates of food insecurity (Niles et al. 2020). In 2012, 12.4% of Canadians households and 11.8% Ontarian households experienced some degree of food insecurity (Gundersen et al. 2018; Tarasuk, Fafard St-Germain, and Mitchell 2019). Recent COVID-19 related food security studies in the US have found a massive increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before (Niles et al. 2020; Wolfson and Leung 2020). Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days(Statistics Canada 2020). Food insecurity is a highly concerning public health issue due the vast health consequences that it can induce. Recent publishing suggests that food insecurity in adults could lead to experiencing more stressful events (El-hajj and Benhin 2021). These increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to health outcomes in populations (Niles et al. 2020)

Data and Methods
===================

>>>>

It is known that the urban food environment, within which people make their daily food choices, is essential in influencing eating behaviours and health outcomes, based on factors such as food availability, ease of geographic accessibility and socio-demographic variations [@paez2010relative; @vanderlee2017food; @widener2018spatial]. However, while there is a wealth of literature that has examined the topic of geographic accessibility to healthy food through the "food desert" concept, there has been little research into accessibility to food banks. Although previous work has explored differences in _accessing_ food banks, such as how some households utilize food banks over short periods of time while others regularly utilize food banks as longer-term resource [e.g., @enns2020experiences], we are not aware of any research that has focused on the geographic component of accessibility. **Check @tarasuk2019geographic**

@@

For this study, we aim to look at how the landscape of food banks and related services (e.g. low-cost or free meal service providers) available in Hamilton, Ontario, has changed before and during the pandemic. Have the number of open food bank services diminished? If so, what was the accessibility to foodbanks before and during the pandemic, from the perspective of low income households? And finally, who are most likely to have been impacted by changes to the accessibility landscape? This paper will first look at the distribution of food banks and related services before and during the pandemic. Then, we use the balanced floating catchment area approach of Paez et al. [@paez2019demand] to investigate the accessibility situation. We use a fully disaggregated approach based on parcel-level data. Socio-economic and demographic data are drawn from the latest Census of Canada (2016), whereas travel information is from the most recent regional travel survey from 2016. This paper follows reproducible research recommendations [see @brunsdon2020opening], and the research was conducted using open source tools for transportation analysis [@lovelace2021open]. The code and data necessary to reproduce the analysis are available in a public repository^[add repository].

@@

Food insecurity is the inability to acquire and consume an adequate amount or good quality food, leading to inadequate nutrient intake [@kirkpatrick2008food; @tarasuk2009household]. This nutrient deficiency has been causing major health concerns in Canadians, and particularly those who are at a socioeconomic disadvantage [@bazerghi2016role]. Previous studies have aimed to look at the relationship between the built food environment and socio-demographic characteristics with qualitative and quantitative, or a mixed -method approach. Quantitatively, official governmental surveys have been able to assist with data, such as the Household Food Security Survey Module (HFSSM), the Canadian Community Health Surveys (CCHS), the Longitudinal and International Study of Adults (LISA), and official classifications determined by Health Canada in relation to  sociodemographic variables [@kirkpatrick2008food; @tarasuk2009household; @gundersen2018food]. Studies have also aimed to assess food availability of healthy foods (e.g., fruit and vegetables) at supermarkets in relation to sociodemographic characteristics and geographic accessibility [@latham2007determinants]. In terms of findings, studies have generated inconsistent relationships between their evaluated availability of food and sociodemographic characteristics that go in hand.

## Foodbanks

Food banks - sometimes also referred to as ‘food pantries’ and ‘food shelves’ - originated as a community response to aid those with inadequate food by voluntarily offering them meals and ingredients [@riches2002food; @loopstra2012relationship]. The scope and objectives of foodbanks can vary by region and by country, and these organizations can include not only prepared meals and aliments, but also shared spaces to connect in community gardens and community kitchens [@wakefield2013sweet]. Although in their origin foodbanks were meant to be provide a temporary solution to accommodate those in hunger due to job retrenchments and economic downfalls since the 1980s, over time they evolved into a community practice to secure emergency food supplies for those in need [@loopstra2012relationship; @wakefield2013sweet]. In Canada, the number of foodbanks has steadily increased in the past few decades [@wakefield2013sweet]. Previous studies have questioned whether food banks are able to offer a rounded nutritious supply of food, and if food banks are a sustainable on-going practice for those in need [@riches2002food; @bazerghi2016role]. An inescapable reality, at the moment, is that despite their imperfections they provide urgently needed support for members of the population for whom foodbanks are frequently their primary location to get food [@bazerghi2016role]. **However, surveys revealed that only 20 to 30 percent of those experiencing food insecurity were found to frequent food banks in Canada [@tarasuk2014food].** **Something about geographical access**

@@

Currently with the COVID-19 pandemic, disrupted economies, rising unemployment rates, and changes in poverty levels have disrupted the food environment and led to higher rates of food insecurity [@niles2020early]. In 2012, 12.4% of Canadians households and 11.8% Ontarian households experienced some degree of food insecurity [@gundersen2018food; @tarasuk2020relationship]. Pandemic-related food security studies in the US have found a substantial increase in households experiencing food insecurity for the first time, and also in households experiencing more severe food insecurity than before [@niles2020early; @wolfson2020food]. Most recently in May 2020, Canada recorded 14.7% of its population living in food insecurity in the past 30 days [@statisticscanada2020food]. Food insecurity is a highly concerning public health issue due to its health impacts. Recent research suggests that food insecurity in adults can lead to experiencing more stressful events (El-hajj and Benhin 2021). Increases in food insecurity rates due to the pandemic, signal a change in the food environment with potential damages to population health during the course of the pandemic and beyond [@niles2020early].

Methods and Materials
===================

## Methods

For the research in this paper we adopt the balanced floating catchment area approach of Paez et al. [-@paez2019demand]. This method for estimating accessibility is a form of the common two-stage floating catchment area method [@radke2000spatial; @luo2003measures]. Floating catchment areas are used to estimate accessibility when there are potential congestion effects, and operate by calculating first the _demand_ for spatially distributed services. The demand (usually the number of people who require a service) is used to calculate a level of service. In a second step, the level of service is allocated back to the population. Demand and level of service are allocated using some form of distance-decay to embody the geographical principle that, given a choice, people prefer to travel less than more when reaching destinations.

More formally, the first step of this method is as follows:
$$
L_j = \frac{S_j}{\sum_{i=1}^nP_iw_{ij}}
$$

\noindent where $S_j$ is the level of supply at location $j$, in simplest terms whether a service point is present (i.e., $S_j=1$) or not (i.e., $S_j=0$); $P_i$ is the population at location $i$ that demands the service; and $w_{ij}$ is a weight, typically a function of the distance between locations $i$ and $j$. $L_j$ is the level of service at location $j$ and it is the inverse of the number of people that need to be serviced.

The second step in this process is then summing the level of service that each population unit can reach, according to the distance-decay weight:
$$
A_i = \sum_{j=1}^JL_jw_{ji}
$$

\noindent where $A_i$ is the accessibility to the service, which is in the same units as the level of service: as the inverse of the population being serviced. When the population being serviced is low accessibility is high (i.e., there is little competition for the service), and viceversa.

Floating catchment area methods are prone to overestimation of the population and the level of service due to multiple-counting. The population at $P_i$ is allocated to _every_ service point $j$ for which $w_{ij}>0$. Similarly, the level of service at $LOS_j$ is allocated to _every_ population point for which $w_{ji}>0$. This inflation effect has been known for several years, and several modifications have been proposed to mitigate it [e.g., @wan2012three; @delamater2013spatial]. A definitive solution to this issue was presented by Paez et al. [-@paez2019demand]. In order to avoid the multiple-counting in the summations, the population and the level of service need to be allocated _proportionally_. This is achieved by standardizing the weights as follows:
$$
w_{ij}^\text{st} = \frac{w_{ij}}{\sum_{i=1}^nw_{ij}}
$$
\noindent and:
$$
w_{ji}^\text{st} = \frac{w_{ji}}{\sum_{j=1}^Jw_{ji}}
$$

The standardized weights satisfy the following conditions:
$$
\sum_{i=1}^nw_{ij}^\text{st}=1
$$

\noindent and:
$$
\sum_{j=1}^Jw_{ji}^\text{st}=1
$$

As a result, the total population (and consequently, the level of service) are preserved system-wide:
$$
\sum_{i=1}^nP_iw_{ij}^\text{st}=P_i
$$

\noindent and therefore:
$$
\sum_{j=1}^JL_jw_{ji}^\text{st}=L_j
$$
